---
title: "*Supplementary Material 2*"
subtitle: "Fossil Calibration List"
mainfont: Roboto-Regular.ttf
fontsize: 11pt
mainfontoptions: 
  - BoldFont=Roboto-Bold.ttf
  - ItalicFont=Roboto-Italic.ttf
  - BoldItalicFont=Roboto-BoldItalic.ttf
output: 
  bookdown::pdf_document2:
    citation_package: default
    toc: no
    number_sections: no
    fig_caption: yes
    keep_tex: no
    latex_engine: xelatex
    pandoc_args: [ "--csl", "frontiers.csl"]
  bookdown::word_document2:
    pandoc_args: [ "--csl", "frontiers.csl"]
    toc: no
    number_sections: no
    reference_docx: Frontiers_Template_SI.docx
header-includes: 
  - \usepackage{float} # keep figures from floating away
  - \usepackage{booktabs} # need for kable tables
  - \usepackage{colortbl} # need for kable tables
  - \makeatletter\renewcommand*{\fps@figure}{H}\makeatother # keep figures from floating away
  - \usepackage{fancyhdr} # enable header
  - \pagestyle{fancy}  # enable header
  - \fancypagestyle{plain}{\pagestyle{fancy}} # show header on first page
  - \renewcommand{\headrulewidth}{0pt} # delete line beneath header
  - \fancyhead[RE,RO]{Supplementary Material} # add header on right side
  - \fancyhead[LO,LE]{} # suppress automatic section header on left side
  - \cfoot{\thepage} # footer
editor_options: 
  chunk_output_type: console
params:
  doc_type: manual
---

---
bibliography: 
- `r here::here("ms/si_fossils.bib")`
---

```{r setup-rmd, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE, results = 'hide')
library(glue)
library(gluedown)
library(tidyverse)
library(assertr)
```

```{r custom-funcs}
#' Print a single category of fossil data
#'
#' @param fossil_data Fossil data in long form, filtered to a single
#' fossil
#' @param cat_select Category of data to print
#'
#' @return Character vector

print_fossil_cat <- function(fossil_data, cat_select) {
  data <-
    fossil_data %>% filter(category == cat_select)
  
  c(
    md_heading(cat_select, 3),
    md_bullet(data$text),
    md_blank()
  )
}

#' Print data for a single fossil
#'
#' @param fossil_data Fossil data in long form
#' @param fossil_select Selected fossil to print
#'
#' @return Character vector
#'
#' @examples
#' print_fossil(fossils_long, "Acrostichum_intertrappeum") %>%
#' 	glue::as_glue()
print_fossil <- function(fossil_data, fossil_select) {
  # Filter data to only selected fossil
  data <-
    fossils_long %>% filter(fossil_taxon == fossil_select)
  
  # Print name of fossil as header in italics
  header <-
    data %>% 
    filter(var == "fossil_taxon_header") %>%
    pull(value) %>%
    str_replace_all("_", " ") %>%
    md_italic() %>%
    md_heading(1)
  
  categories <- data %>%
    filter(!is.na(category)) %>%
    pull(category) %>%
    unique()
  
  body <- map(categories, ~print_fossil_cat(data, .)) %>%
    unlist()
  
  c(header, body, md_rule())
}
```

```{r prep-fossil-data}
# Set path to fossil data
fossils_data_path <- here::here("_targets/user/data_raw/Fossils_Ferns.csv")

# Extract fossil categories (first row of fossil data)
fossils_categories <-
  read_csv(fossils_data_path, n_max = 1) %>%
  pivot_longer(names_to = "category", values_to = "raw_names", everything()) %>%
  mutate(
    category = str_replace_all(category, "\\.\\.\\.[0-9]+", "skip") %>%
      na_if("skip")) %>%
  fill(category) %>%
  mutate(category = str_to_title(category)) %>%
  filter(category != "Pteridophy")

# Get raw column names
fossils_raw_names <-
  read_csv(fossils_data_path, skip = 1)

# Load data with cleaned column names
fossils_clean_names <- fossils_raw_names %>% janitor::clean_names()

# Tidy fossils in wide format
fossils <-
  fossils_clean_names %>%
  # Filter to only those actually used in the study
  filter(used == "X") %>%
  select(-used) %>%
  select(fossil_taxon, everything())

# Make a tibble of column namees for formatting
name_tbl <- tibble(
  raw_names = colnames(fossils_raw_names),
  var = colnames(fossils_clean_names)
) %>%
  right_join(
    tibble(var = colnames(fossils)), by = "var"
  ) %>%
  mutate(var_order = 1:nrow(.))

# Convert to long format
fossils_long <- fossils %>%
  mutate(
    across(everything(), as.character),
    fossil_taxon_header = glue("†{fossil_taxon}")) %>%
  pivot_longer(names_to = "var", values_to = "value", -fossil_taxon) %>%
  left_join(name_tbl, by = "var") %>%
  mutate(
    text = as.character(glue::glue("{raw_names}: {value}"))
  ) %>%
  left_join(fossils_categories, by = "raw_names") %>%
  arrange(fossil_taxon, category, var_order)
```

These data are exported from the FernCal v1.0 database (https://github.com/pteridophy/FossilFerns), which includes `r nrow(fossils_clean_names)` records of fossils to use as calibration points for molecular dating of ferns.

"NFos" is the unique ID of the record in FernCal.

Taxonomy of extant taxa follows pteridocat v1.0 database (https://github.com/pteridophy/pteridocat).

For a summary of these data in CSV format, see Table FIXME <add table number>.

\newpage

```{r print-fossils, include = TRUE, results = "asis", eval = TRUE}
unique(fossils_long$fossil_taxon) %>%
  map(~print_fossil(fossils_long, .)) %>%
  unlist() %>%
  glue::as_glue()
```

# References